Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)

### 2024-11-27 paper/paper.qmd
introduction

# load data
cleaned_expectancy <- read_parquet(here:here("data/02-analysis_data/analysis_data.parquet"))

@@

# Introduction
Life expectancy is a critical measure of a population's overall health and well-being, shaped by various factors such as personal health, government healthcare access, economic stability, and social structures. The index of life expectancy is generally served as a benchmark for development, indicating the effectiveness of interventions in reducing mortality and improving well-being. Higher life expectancy is believed to linked to better living standards, access to education, and equitable healthcare. Conversely, low life expectancy often signals systemic challenges such as poverty, disease burden, and inadequate healthcare access. Policymakers and researchers rely on life expectancy data to design targeted strategies, allocate resources, and monitor progress toward achieving global health and development goals.

This study focuses on understanding the relationship between country status(developed or developing), personal average expenditure on health based on GDP, general BMI of entire population, percentage of government expenditure on health, GDP, income composition of resources and year of schooling in predicting life expectancy across different countries. By identifying and analyzing these predictors, the study aims to highlight actionable aspects for policymakers to target in their efforts to improve population longevity effectively.

@@

>>>>

# load data
cleaned_expectancy <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))

@@

# Introduction
Life expectancy is a critical measure of a population's overall health and well-being, shaped by various factors such as personal health, government healthcare access, economic stability, and social structures. The index of life expectancy is generally served as a benchmark for development, indicating the effectiveness of interventions in reducing mortality and improving well-being. Higher life expectancy is believed to linked to better living standards, access to education, and equitable healthcare. Conversely, low life expectancy often signals systemic challenges such as poverty, disease burden, and inadequate healthcare access [@socialdeterminantsofhealth].

The primary goal of this paper is to determine which factors play a statistically significant role in driving lower life expectancy values and to offer actionable insights based on country status. Using Multiple Linear Regression model, this study focuses on understanding the relationship between country status(developed or developing), personal average expenditure on health based on GDP, general BMI of entire population, percentage of government expenditure on health, GDP, income composition of resources and year of schooling in predicting life expectancy across different countries. By identifying and analyzing these predictors, the study aims to highlight actionable aspects for policymakers to target in their efforts to improve population longevity effectively.

@@

Findings from this paper reveal that countries with greater healthcare spending tend to achieve higher life expectancy, while factors like income inequality and absence of education emerge as significant obstacles. The analysis underscores the importance of targeted interventions in key areas such as healthcare accessibility and economic equity to address disparities in life expectancy. This study further contributes to a deeper understanding of how predictive modeling under different country status can inform public health strategies and policy-making on a global scale.
